The ``Dark Side'' of Phylogenetic Comparative Methods

### 2014-08-28 Manuscript/dark.tex
completed draft of problems section

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (``wookies'') develop new methods and implement them. They are (generally) aware of the limitations of their methods, and the assumptions that underly them. 
% Bit of a stretch but we can pretend right? :)
At the other extreme, na\"{i}ve methods-users (``ewoks'') just want to use PCMs to answer empirical questions with their data, and may not be aware of issues with the methods. 
Of course the majority of people fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use. 
The problem is that although methods-developers (and advanced methods-users) are aware of the problems with PCMs, this information is not being effectively transferred to methods-users. Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. Here we explore the causes of this communication gap and suggest some potential solutions. %still not happy with this writing

@@

As scientists we mainly communicate our ideas and through the literature. Therefore, any difficulty transferring information between methods-developers and methods-users must begin with the literature. If this is the case we suggest three possible reasons:
%Yuk fix this

@@

A lot of the information needed to properly apply PCMs is not found in the literature. 
We refer to this knowledge as ``PCM folklore'' because it tends to be passed down from PIs to graduate students.
Sometimes the folklore includes tricks to get methods working, or useful rules-of-thumb. 
Other times folklore is more opinion-based, but over time these myths become set in stone. 
One example of such folklore is in the defaults of programs that perform PCMs. 
These often start out as arbitrary starting points for data exploration with no justification for their use, but over time become the way the analysis is always performed.
Useful PCM folklore is often shared among methods-developers, and among collaborating groups, but is rarely shared outside of these circles. 
When it is shared, it tends to be in the form of email exchanges between methods-developers and struggling methods-users.
This advice is extremely useful to the methods-user in question, but forms a kind of ``dark advice'' that isn't accessible to the rest of the community.\\

@@

Other information about the limitations of a method may be absent from the literature due to the time lag between a new method being published and others having time to test and critique it. 
For example, \citet{felsenstein1985phylogenies} published the phylogenetic independent contrasts method in 1985, but it wasn't until the early 1990s that real critiques of the method and its assumptions began to be published. %add citations.
This time lag is shorter with more recent methods because simulations testing the method are now required by journals.
However, we suspect there are still many hidden assumptions and biases in virtually all PCMs.

@@

More recently however, the PCM community has moved towards mostly implementing their methods in R (cite), and code sharing has become almost ubiquitous. 
The number of R packages for PCMs has increased exponentially since 2005 when ape (A=P=E=, cite) was released, and has increased particularly sharply since 2008 (Figure \ref{x}).
Simultaneously, more and more methods-users are able to use R, thus when a new method is published they can now take an R package ``off the shelf'' and use it to run the method immediately.
We are all big proponents of Open Science so we think this is a wonderful thing for the field, and we do not mean to criticize 
However, this has led to an increasing number of methods-users jumping straight to the implementation of a method, without really carefully trying to understand what the method is doing, or what it's assumptions might be. 
Additionally, many methods-users seem to try to retrofit a question so they can use the newest shiniest methods available. 
All of us regularly review papers where the authors have clearly not read the original papers very carefully.

This problem is exacerbated by the fact that manuals, vignettes and help files for R packages rarely mention the assumptions of the method, or how to test model fit

%This increase in use is at least partially the result of the number of available R packages to run these analyses (for example ape, GEIGER, diversitree, caper - citations) and the community's increasing familiarity with R \citep{R-Core-Team:2014aa}. Unfortunately with this increase in popularity, comes an increase in misuse! % need to work on this paragraph

%Huge increase in number of PCM packages since 1980s, and especially since 2008 [Graph]. Now people familiar with R they can just jump in and try stuff on their data with very little knowledge of what is going on or what the results mean [Though code sharing is cool and I’m not dissing that].

%I’m torn as to what I think here – part of me thinks there should be a mention of assumptions in the manual/vignette (though not if this replicates an accompanying blog or book or something) but part of me thinks it’s the users responsibility to not be a fuckwit. Caper, diversitree and CAIC are good examples where assumptions are clearly laid out and methods for testing assumptions are available in caper and CAIC (not sure about diversitree ...). Ape book has none of this in it at all!

@@

\subsection{1. Simplify, summarize and share}
\subsection{2. Read, think and learn}
\subsection{3. Collaboration}
\subsection{4. Incentives}

%Such issues are the responsibility of end users but also of methods developers: the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. To address these issues, we propose this symposium to discuss issues in both classical and recent PCMs, along with new research on detecting for these issues and accounting for them.  We hope that this will both increase awareness of these problems and encourage further research and careful thought in the area, along with better dialogue between method developers and method users.

%Big issue = who is responsible? Can't expect empiricists to be an expert in every method, but do want to add to workload of overloaded developers.

%Incentives - funding for pure methods. Different authorship rules.

%Change the definition of success in academia. Quality not quantity. Remove rush to publish. This would help WOMEN too!

%Add methods to test assumptions as standard to packages.

@@

@@

>>>>

One major issue is that two groups of people are involved with PCMs. 
At one extreme, methods-developers (``wookies'') develop new methods and implement them. They are (generally) aware of the limitations of their methods, and the assumptions that underly them. 
% Bit of a stretch but we can pretend right? :)
At the other extreme, na\"{i}ve methods-users (``ewoks'') just want to use PCMs to answer empirical questions with their data, and may not be aware of issues with the methods 
%when the next sentence is not in brackets it breaks the flow. But needs to be here. Brackets feel wrong too. Will Syst Biol allow footnotes
(of course the majority of people fall somewhere between these two extremes on the ``wookie-ewok'' spectrum; many methods-developers are developing methods with a specific empirical question in mind, while there are many advanced methods-users who have detailed knowledge of the limitations of the methods they use). 
The problem is that although methods-developers (and advanced methods-users) are aware of the problems with PCMs, this information is not being effectively transferred to methods-users. 
Additionally, the tools and approaches used to fit models are often far more user-friendly and better documented than the methods used to to assess whether that model fit is reasonable. 
Clearly more effort is needed to bridge the widening gap between methods-users and methods-developers. Here we explore the causes of this communication gap and suggest some potential solutions. %still not happy with this writing

@@

As scientists we mainly communicate our ideas through the literature. Therefore, any difficulty transferring information between methods-developers and methods-users must begin there. If this is the case we suggest three possible reasons:
%Yuk fix this

@@

A lot of the information needed to properly apply PCMs is not found in the literature. 
We refer to this knowledge as ``PCM folklore'' because it tends to be passed down from PIs to graduate students, and it is occasionally closer to fiction than fact.
Sometimes the folklore includes tricks to get methods working, or useful rules-of-thumb. 
Other times folklore is more opinion-based, but over time these opinions become rules. 
One example of such folklore is in the defaults of programs that perform PCMs. 
These often start out as arbitrary starting points for data exploration with no justification for their use, but over time become the way the analysis is always performed.
Useful PCM folklore is often shared among methods-developers, and among collaborating groups, but is rarely shared outside of these circles. 
When it is shared, it tends to be in the form of email exchanges between methods-developers and struggling methods-users.
This advice is extremely useful to the methods-user in question, but forms a kind of ``dark advice'' that isn't accessible to the rest of the community.\\

@@

Other information about the limitations of a method may be absent from the literature due to the time lag between a new method being published and others having time to test and critique it. 
For example, Felsenstein published the phylogenetic independent contrasts method in 1985, but it wasn't until the early 1990s that real critiques of the method and its assumptions began to be published. %add citations.
This time lag is shorter with more recent methods because simulations testing the method are now required by journals.
However, we suspect there are still many hidden assumptions and biases in virtually all new PCMs that have yet to be properly explored in the literature. 
%We can cite your paper on Pagel 1994 here if you like - as an example of old methods that are still being investigated.

@@

More recently however, the PCM community has moved towards mostly implementing their methods in R (cite), and code sharing has become almost ubiquitous. 
The number of R packages for PCMs has increased exponentially since 2005 when APE (cite) was released, and has increased particularly sharply since 2008 (Figure \ref{x}).
Simultaneously, more and more methods-users are able to use R, thus when a new method is published they can now take an R package ``off the shelf'' and use it to run the method immediately.
This has led to an increasing number of methods-users jumping straight to the implementation of a method, without really carefully trying to understand what the method is doing, what its assumptions might be, or what the results they extract really mean in a biological context.
This problem is exacerbated by the fact that manuals, vignettes and help files for R packages rarely mention the assumptions of the method, or how to test model fit. 
The APE book (cite) for example, provides no guidance on assumption testing in its chapter on phylogenetic independent contrasts, even though the methods needed to do this are well-established and very easily implemented in R (for a counter example see CAIC and caper documentation - cite).\\
We believe this issue is probably one of the biggest currently facing phylogenetic comparative biology. It is extremely frustrating reading/reviewing papers where the authors have clearly not carefully read or understood the primary literature on the method they are using. 
Additionally, many methods-users seem to try to retrofit a question so they can use the newest shiniest methods available. This is all driven by our ``publish or perish'' culture, and the pressure to produce as many papers as possible out of a dataset.
The truly sad thing about this state of affairs is that by making science more open and sharing code (something we are all passionate about and strongly support) we seem to have made the situation worse rather than better (\citealp{freckleton2009seven}, for example, mentions R as a way of improving things). %needs a bit of finessing

@@

The problems described above require action by methods-developers, methods-users and the community as a whole. %Can't expect empiricists to be an expert in every method, but do want to add to workload of overloaded developers.

\subsection{1. Simplify, summarize and share}

a) dealing with

@@

May not be doable... List too many assumptions...
Discourse pages – forums for particular things. Dark “advice”.
Reduces time needed. 
Phylobabble

\subsection{2. Read, think and learn}

@@

\subsection{3. Collaboration}

A obvious solution to all these problems is for methods-developers and methods-users to collaborate more. The benefits to methods-users is obvious, but methods-developers can gain extra datasets to test their ideas on, people who will discover corner cases and bugs in their software before it get released, and gain an appreciation of . 
A big difficulty with collaboration

\subsection{4. Incentives}

%Incentives - funding for pure methods. Different authorship rules. 
%Change the definition of success in academia. Quality not quantity. Remove rush to publish. This would help WOMEN too!

%Add methods to test assumptions as standard to packages. 
%I’m torn as to what I think here – part of me thinks there should be a mention of assumptions in the manual/vignette (though not if this replicates an accompanying blog or book or something) but part of me thinks it’s the users responsibility to not be a fuckwit. Caper, diversitree and CAIC are good examples where assumptions are clearly laid out and methods for testing assumptions are available in caper and CAIC (not sure about diversitree ...). Ape book has none of this in it at all!

Encourage quality over quantity % this is how I'm apparently allowed to justify my publication record
Additionally, many methods-users seem to try to retrofit a question so they can use the newest shiniest methods available.
